Musical elements in the discrete-time representation of sound

### 2017-07-29 article/acm/body.tex
revisada parte de ritmo

\subsection{Counterpoint}\label{subsec:contraponto}

Counterpoint is defined as the conduction of simultaneous melodic lines, or ``voices''. The bibliography covers systematic ways to conduct voices, leading to scholastic genres like canons, inventions and fugues. It is possible to
summarize counterpoint rules, and it is known that Beethoven --
among others -- also outlined such a briefing of counterpoint.

@@

The purpose of counterpoint is to conduct voices in a way that they sound independent. In order to do that, the relative motion of voices (in pairs) is crucial and
categorized as: direct, oblique and contrary (see Figure~\ref{fig:movContraponto}).
The parallel motion is a direct motion.
The gold rule here is to take care of the direct motions, avoiding them
when leading to a perfect consonance. The parallel motion should occur only between
imperfect consonances and no more than three consecutive times. Disonances can be unadmitted or used when followed and preceded by consonances of joint
degrees, i.e.\ neighbor notes in a scale. The motions that lead to a
neighbor note in the scale sound coherent. When having 3 or more voices, the melodic
importance lies in the higher and lower voices, in this order~\cite{Fux,Tragtenberg,SchoenbergContra}.

@@

\subsection{Rhythm}\label{subsec:ritmo}

Rhythmic notion is dependent on events separated by durations~\cite{Lacerda}, which can be heard individually if spaced by at least $50-63ms$. For the temporal separation between them to be perceived as duration, the period should even a bit larger, around $100ms$~\cite{microsound}.  
It is possible to summarize duration heard as rhythm or pitch, and its transition,
as in table~\ref{tab:duracoes}~\cite{Alfaix, microsound}.

\begin{table*}[htp!]
\caption{Transition of durations heard as rhythm until heard as pitch.}
\begin{tabular}{  l | r r r r   r r r    r r r || r r  } \hline
& \multicolumn{10}{c}{\bf perception of durations as rhythm} & \multicolumn{2}{c}{}  \\
	duration (s) & \multirow{2}{*}{\bf ...}     & {\bf 32,}     & {\bf 16,}   & {\bf 8,}  & {\bf 4,}   & {\bf 2,}   & {\bf 1,}   & {\bf 1/2,} & {\bf 1/4,} & {\bf 1/8,} & \multirow{2}{*}{ ... transition } & \\
frequency (Hz) & & {\color{gray} 1/32,}   & {\color{gray} 1/16,} & {\color{gray} 1/8,} & {\color{gray} 1/4,} & {\color{gray} 1/2,} &  {\color{gray} 1,}  & {\color{gray} 2,}   & {\color{gray} 4,}   & {\color{gray} 8,} & & \\
& \multicolumn{10}{c}{ - } & \multicolumn{2}{c}{} \\ \hline
\end{tabular}
\vspace{.2cm}

@@

The duration transition span is minimized because the limits
are not well defined. In fact, the duration where someone begins to perceive a
fundamental frequency or a separation between occurrences, depends on the
listener and sonic characteristics~\cite{microsound,Roederer}.

The rhythmic metric is commonly based on a basic duration called pulse, whose durations range between $0.25 and 1.5s$ ($240$
and $40 BPM$, respectively). In music education and cognitive studies, it is common to associate these range of frequencies to the durations of the heart beat, movements of inspiration and expiration and steps of a walking or running person~\cite{Lacerda,Roederer}.

The pulse is subdivided into equal parts and is also repeated in sequence. These relations (division and concatenation) usually follow relations of small
order integers. The occurrences, in written and ethnic
music, of musical pulse division (and their sequential grouping at
time), are, in ascending order: 2, 4 and 8, after that 3, 6 (two groups of 3 or 3 groups of 2) and 9 and 12 (three and 4 groups of 3). At last, the prime numbers 5 and 7, complementing
1-9 and 12. Other metrics are less common, like division and grouping in 13, 17, etc, and are mainly used in contexts of experimental music and classical music of XX and XXI. No matter how complex they seem, metrics are common compositions and decompositions of 1-9 equal parts~\cite{Gramani,Roederer}.
A schematic illustration is shown in Figure~\ref{fig:pulsoSubAgl}.

@@

Dual relations (simple measures and binary divisions) are frequent in dance rhythms and celebrations, and are called ``imperfect''. Ternary relations are common in
ritualistic music and is related to the sacred, and are called ``perfect''. Strong units (accents) fall in the `head' of bars, the downbeats, of which the first in
every bar is the greatest. The head of a unit is the first part of the subdivision. In binary divisions (2, 4
and 8), strong units alternate with weak units
(e.g.\ division in 4 is: strong, weak, average strong, weak). In ternary divisions
(3, 6 and 9) two weak units succeed the downbeat (e.g.\ division in 3 is:
strong, weak, weak). Division in 6 is considered compound but can also
occur as a binary division. Binary division units which suffers a ternary division yields two units divided into tree units each: strong (subdivided in strong,
weak, weak) and weak (subdivided in strong, weak, weak). Another way to perform
the division in 6 is ternary division units that divides as binary,
resulting in: a strong unit (subdivided in strong and weak) and two weak units
(subdivided in strong and weak for each).

The accent in the weak beat is the backbeat, whereas a note starting in a weak beat persisting across the strong beat is the syncope.

Notes can occur inside and outside of these \emph{'musical metric'} divisions. In most well-behaved cases, notes occur exactly on these divisions, with greater incidence on strong beat attacks.
In extreme cases, time metric cannot be perceived~\cite{Roederer}. 
Small variations along the grid are crucial for musical interpretation and
different styles~\cite{Cook}.

Let the pulse be the grouping level $j=0$, the first pulse subdivision  be level $j=-1$,
the first pulse agglomeration be level $j=1$ and so on. Hence, $P_i^j$ is the $i$-th unit at grouping level $j$: $P^0_{10}$ is the tenth pulse, $P^{1}_3$ is the third grouped unit (and, possibly, the third measure),
$P^{-1}_2$ is the second part of pulse subdivision. The limits of $j$ are of special interest: pulse divisions are durations perceivable as rhythm; furthermore, the pulses sum, at its maximum, a music or a cohesive set of musics. In other words, a duration given
by $P^{min(j)}_i$, $\forall \; i$, is greater than $50 ms$ and the durations
summed together $\sum_{\forall i}P^{\text{max}(j)}_i$ are less than a few
minutes or, at most, a few hours.

Each level $j$ has some sections $i$. When $i$ has three different
values (or multiple of three) there is a perfect relation. When $i$ has only
two, four or eight different values, than there is an imperfect relation,
as shown in Figure~\ref{fig:pulsoSubAgl}. Any unit (note) of a given musical sequence with a time metric can be unequivocally
specified as:

@@

\noindent where $j_k$ is the grouping level and$i_k$ is the unit itself.

As an example, consider $P^{-1,0,1}_{3,2,2}$ as the third subdivision $P^{-1}_3$ of the
second pulse $P^0_2$ and of the second pulse group $P^1_2$ (possibly second measure). Each unit $P_i^j$ can be associated with a sequence of temporal samples $T_i$ that comprehends a
note. The music piece \emph{Poli Hit Mia} uses different metrics (available as part of \massa\ toolbox).

\subsection{Repetition and variation: motifs and larger units}\label{subsec:motivos}

Given the basic musical structures, both frequential (chords and scales) and rhythmic (simple, compound and complex beat divisions and agglomerations), it is
natural to present these structures in a coherent and meaningful way~\cite{Boulez}. The concept of arcs is essential in this context: by departing from a place and returning, an arc is made. The audition of melodic and harmonic lines is permeated by
musical arcs due to the cognitive nature of the musical hearing. 
The note can be considered the smaller arc, and each motif and melody as an arc as well.
Each beat and subdivision, each measure and music
section, constitutes an arc. A music in which the arcs do not present consistency with one another,
can be understood as a music with no coherence. Coherence impression
comes, mostly, from the skilled handling of arcs in a music piece.

>>>>

\subsection{Counterpoint}\label{subsec:contraponto}
Counterpoint is the conduction of simultaneous melodic lines, or ``voices''. The bibliography covers systematic ways to conduct voices, leading to scholastic genres like canons, inventions and fugues~\cite{count1, count2}. It is possible to
summarize counterpoint rules, and it is known that Beethoven --
among others -- also outlined such a digest of counterpoint.

@@

The purpose of (scholastic) counterpoint is to conduct voices in a way that they sound independent. In order to do that, the relative motion of voices (in pairs) is crucial and
categorized as: direct, oblique and contrary, as depicted in Figure~\ref{fig:movContraponto}.
The parallel motion is a direct motion in which the starting and final intervals are the same.
The golden rule here is to take care of the direct motions, avoiding them
when ending in a perfect consonance. The parallel motion should occur only between
imperfect consonances and no more than three consecutive times. Dissonances can be forbidden or used when followed and preceded by consonances of neighbor
degrees, i.e.\ adjacent notes in a scale. The motions that lead to a
neighbor note in the scale sound coherent and are prioritized. When having 3 or more voices, the melodic
importance lies mainly in the highest and then in the lowest of the voices~\cite{Fux,Tragtenberg,SchoenbergContra}.

@@

\subsection{Rhythm}\label{subsec:ritmo}
Rhythmic notion is dependent on events separated by durations~\cite{Lacerda}, which can be heard individually if spaced by at least $50-63ms$. For the temporal separation between them to be perceived as duration, the period should even a bit larger, around $100ms$~\cite{microsound}.
It is possible to summarize the durations heard as rhythm or pitch, and its transition,
as in table~\ref{tab:duracoes}~\cite{Alfaix, microsound}.

\begin{table*}[htp!]
\caption{Durations heard as rhythm, as pitch and transition.}
\begin{tabular}{  l | r r r r   r r r    r r r || r r  } \hline
& \multicolumn{10}{c}{\bf perception of durations as rhythm} & \multicolumn{2}{c}{}  \\
	duration (s) & \multirow{2}{*}{\bf ...}     & {\bf 32,}     & {\bf 16,}   & {\bf 8,}  & {\bf 4,}   & {\bf 2,}   & {\bf 1,}   & {\bf 1/2,} & {\bf 1/4,} & {\bf 1/8,} & \multirow{2}{*}{ ... transition } & \\
frequency (Hz) & & {\color{gray} 1/32,}   & {\color{gray} 1/16,} & {\color{gray} 1/8,} & {\color{gray} 1/4,} & {\color{gray} 1/2,} &  {\color{gray} 1,}  & {\color{gray} 2,}   & {\color{gray} 4,}   & {\color{gray} 8,} & & \\
& \multicolumn{10}{c}{ - } & \multicolumn{2}{c}{} \\ \hline
\end{tabular}
\vspace{.2cm}

@@

The transition span in Table~\ref{tab:duracoes} is minimized because the limits
are not well defined. In fact, the duration where someone begins to perceive a
fundamental frequency or a separation between occurrences, depends on the
listener and sonic characteristics~\cite{microsound,Roederer}. The rhythmic metric is commonly based on a basic duration called pulse, whose typical durations range between $0.25 and 1.5s$ ($240$
and $40 BPM$, respectively~\footnote{BPM stands for Beats Per Minute and is just a frequency measure like Herz, but is the number of incidences per minute intead of second. BPM is often used as a measure of musical tempo and of heart rate.}). In music education and cognitive studies, it is common to associate this range of frequencies with the durations of the heart beat, movements of respiration and steps of a walking or running person~\cite{Lacerda,Roederer}.

The pulse is subdivided into equal parts and is also repeated in sequence. These relations (division and concatenation) usually follow relations of small
integers. By far, the most often musical pulse divisions (and their sequential groupings), in written and ethnic
music, are: 2, 4 and 8; 3, 6 (two groups of 3 or 3 groups of 2), 9 and 12 (three and 4 groups of 3). At last, the prime numbers 5 and 7, completing
1-9 and 12. Other metrics are less common, like division or grouping in 13, 17, etc, and are mainly used in experimental music or classical music of the XX and XXI centuries. No matter how complex they seem, metrics are almost always compositions and decompositions of 1-9 equal parts~\cite{Gramani,Roederer}.
A schematic illustration is shown in Figure~\ref{fig:pulsoSubAgl}.

@@

Binary divisions are frequent in dance rhythms and celebrations, and are called ``imperfect''. Ternary relations are common in
ritualistic music and is related to the sacred, and are called ``perfect''. Strong units (accents) fall in the `head' of the units (the first subdivision) and are called downbeats. In binary divisions (2, 4
and 8), strong units alternate with weak units
(e.g.\ division in 4 is: strong, weak, average strong, weak). In ternary divisions
(3, 6 and 9) two weak units succeed the downbeat (e.g.\ division in 3 is:
strong, weak, weak). Division in 6 is considered compound but can also
occur as a binary division. Binary division units which suffer a ternary division yields two units divided into three units each: strong (subdivided in strong,
weak, weak) and weak (also subdivided in strong, weak, weak). Another way to perform
the division in 6 is ternary division units that subdivides as binary,
resulting in: a strong unit (subdivided in strong and weak) and two weak units
(subdivided in strong and weak each).

An accent in the weak beat is the `backbeat', whereas a note starting in a weak beat persisting across the strong beat is a `syncope'.
These are often found in ethnic and popular music and was used with parsimony in classical music before the XX century.

Notes can occur inside and outside of these \emph{'musical metric'} divisions. In most well-behaved cases, notes occur exactly on these divisions, with greater incidence on strong beats.
In extreme cases, rythmic metric cannot be perceived~\cite{Roederer}. 
Small variations along the temporal grid are crucial for musical interpretation and
styles~\cite{Cook}.

Let the pulse be the grouping level $j=0$, the first pulse subdivision be level $j=-1$,
the first pulse agglomeration be level $j=1$ and so on. Accordingly, let $P_i^j$ be the $i$-th unit at grouping level $j$: $P^0_{10}$ is the tenth pulse, $P^{1}_3$ is the third grouped unit (possibly the third measure),
$P^{-1}_2$ is the second part of pulse subdivision. The limits of $j$ are of special interest: pulse divisions are durations perceivable as rhythm; furthermore, the pulses sum, at its maximum, a music or a cohesive set of musical pieces. In other words, a duration given
by $P^{min(j)}_i$, $\forall \; i$, should be greater than $50 ms$ and the durations
summed together $\sum_{\forall i}P^{\text{max}(j)}_i$ should be less than a few
minutes or, at most, a few hours.

Each level $j$ has some sections $i$. When $i$ has three different
values (or multiple of three) there is a perfect (i.e. ternary) relation. When $i$ has only
two, four or eight possible values, than there is an imperfect relation,
as shown in Figure~\ref{fig:pulsoSubAgl}. Any unit (note) of a given musical sequence with a time metric can be unequivocally
specified as:

@@

\noindent where $j_k$ is the grouping level and $i_k$ is the unit itself.

As an example, consider $P^{-1,0,1}_{3,2,2}$ as the third subdivision $P^{-1}_3$ of the
second pulse $P^0_2$ and of the second pulse group $P^1_2$ (possibly second measure). Each unit $P_i^j$ can be associated with a sequence of temporal samples $T_i$ that constitutes a
note. The music piece \emph{Poli Hit Mia} uses different metrics (available as part of \massa\ toolbox).

\subsection{Repetition and variation: motifs and larger units}\label{subsec:motivos}
Given the basic musical structures, both frequential (chords and scales) and rhythmic (simple, compound and complex beat divisions and agglomerations), it is
natural to present these structures in a coherent and meaningful way~\cite{Boulez}. The concept of arcs is essential in this context: by departing from a place and returning, an arc is made. The audition of melodic and harmonic lines is permeated by
musical arcs due to the cognitive nature of the musical hearing. 
The note can be considered the smaller arc, and each motif and melody as an arc as well.
Each beat and subdivision, each measure and music
section, constitutes an arc. A music in which the arcs do not present consistency with one another,
can be understood as a music with no coherence. Coherence impression
comes, mostly, from the skilled handling of arcs in a music piece.

### 2017-07-29 article/acm/foobody.tex
revisada parte de ritmo

\subsection{Counterpoint}\label{subsec:contraponto}

Counterpoint is defined as the conduction of simultaneous melodic lines, or ``voices''. The bibliography covers systematic ways to conduct voices, leading to scholastic genres like canons, inventions and fugues. It is possible to
summarize counterpoint rules, and it is known that Beethoven --
among others -- also outlined such a briefing of counterpoint.

@@

\subsection{Rhythm}\label{subsec:ritmo}

Rhythmic notion is dependent on events separated by durations~\cite{Lacerda}, which can be heard individually if spaced by at least $50-63ms$. For the temporal separation between them to be perceived as duration, the period should even be a somewhat larger, around $100ms$~\cite{microsound}.  
It is possible to summarize duration heard as rhythm or pitch, and its transition,
as in table~\ref{tab:duracoes}~\cite{Alfaix, microsound}.

>>>>

\subsection{Counterpoint}\label{subsec:contraponto}
Counterpoint is defined as the conduction of simultaneous melodic lines, or ``voices''. The bibliography covers systematic ways to conduct voices, leading to scholastic genres like canons, inventions and fugues. It is possible to
summarize counterpoint rules, and it is known that Beethoven --
among others -- also outlined such a briefing of counterpoint.

@@

\subsection{Rhythm}\label{subsec:ritmo}
Rhythmic notion is dependent on events separated by durations~\cite{Lacerda}, which can be heard individually if spaced by at least $50-63ms$. For the temporal separation between them to be perceived as duration, the period should even be a somewhat larger, around $100ms$~\cite{microsound}.  
It is possible to summarize duration heard as rhythm or pitch, and its transition,
as in table~\ref{tab:duracoes}~\cite{Alfaix, microsound}.
